Factors determining how catchments respond to forest cover change. Re-analysing global data sets.

### 2022-01-12 paper/Forest_and_Water/Forest_and_Water.Rmd
moving towards a final version'

---
title: Factors determining how catchments respond to forest cover change: Re-analysing global data sets.
author:
  - name: R. Willem Vervoort
    email: willem.vervoort@sydney.edu.au
    affiliation: DARE, The University of Sydney
    footnote: 1
  - name: Eliana Nervi
    email: eliananervi@gmail.com
    affiliation: INIA
  - name: Jimena Alonso
    email: jalonso@fing.edu.uy
    affiliation: IMFIA
address:
  - code: DARE
    address: ARC Training Centre Data Analytics for Resources and the Environment
  - code: The University of Sydney
    address: School of Life and Environmental Sciences, The University of Sydney, Sydney, NSW 2006, Australia
  - code: INIA
    address: Project Manager, FPTA 357, Instituto Nacional de Investigación Agropecuaria, INIA-Uruguay, Ruta 48 km 10, Rincon del Colorado, 90100 Canelones, Uruguay
  - code: IMFIA
    address: Institute of Fluid Mechanics and Environmental Engineering, School of Engineering, Universidad de la República, 11200 Montevideo, Uruguay
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: |
  Three recent papers review and analyse large global datasets related to impacts of forest cover on streamflow. Using three different approaches, they all find a strong relationship between forestation/de-forestation and streamflow. However, the past approaches in the literature are variable and can be substantially improved in statistical rigour, and indicate different confounding factors on the impact of forestation. The data for these three papers were reviewed, combined and re-analysed to answer the following new and older questions: 1) How is streamflow impacted by the change in forest cover as a function of catchment area; 2) how is this relationship conditioned by the length of the study, and climate; and 3) are there other possible variables that impact the observed change in streamflow? Generalised additive models were used to run flexible regressions including multiple variables. 
  Changes in forest cover cause changes in streamflow, however this change is different between deforestation and reforestation, and strongly affected by climate, with drier climates indicating larger changes in streamflow. Deforestation causes a 32\% greater change in flow compared to reforestation. Area of the catchment only affects the change in streamflow after log transformation, given the wide variety in the data from small scale paired watershed studies. Smaller studies dominate the database with 42\% of the data < 1 km2 and 65\% of the data < 10 km2. Length of the study and initial year of the study did not affect the change in flow, in contrast to other reported studies. Despite these findings, overall explained variance (38/%) is low due the quality of the inputs and additional unknown confounding factors.
journal: "Journal of Hydrology"
date: "`r Sys.Date()`"
bibliography: forest_and_water.bib
linenumbers: true
#numbersections: true
csl: elsevier-harvard.csl
output:   
  bookdown::pdf_book:
    base_format: rticles::elsevier_article
editor_options: 
  chunk_output_type: inline
---

@@

<!-- Include Filoso paper -->

@@

The conclusions of the first paper [@zhang2017] suggest that there is a distinct difference in the change in flow as a result of forestation or deforestation between small watersheds, defined as < 1000 km^2^ and large watersheds > 1000 km^2^. While for small watersheds there was no real change in runoff with changes in cover, for large watersheds there was a clear trend showing a decrease in runoff with and increase in forest cover. Their main conclusion was that the response in annual runoff to forest cover was scale dependent and appeared to be more sensitive to forest cover change in water limited watersheds relative to energy limited watershed [@zhang2017].

The second study [@filoso2017] was a systematic review which classified the historical research and highlighted gaps in the spatial distribution, the types of studies and the types of analysis. Their main conclusion was also that reforestation decreases streamflow, but that there were many interacting factors. For a subset of quantitative data (37) they showed a log-linear relationship between decreasing watershed size and an increasing decline in streamflow. In addition, they identified that studies with shorter periods of data collection resulted in larger declines in streamflow.

@@

The starting point of this paper is the data base of studies which were included in @zhang2017 as supplementary material. The columns in this data set are the watershed number, the	watershed name, the	Area in km^2^, the annual average precipitation	(Pa) in mm, the	forest type, hydrological regime, and climate type, the change in forest cover	in \% ($\Delta F\%$) and the change in streamflow in \% $(\Delta Qf\%)$, based on equation 1 in @zhang2017), the precipitation data type, the assessment technique, and the	source of the info, which is a citation. 
Several of these columns contain abbreviations to describe the different variables, which are summarised in Table 1.

@@

While @zhang2017 use the dryness index in their analysis, potential or reference evapotranspiration was not originally included as part of the published data set.
We combined the tables for small (< 1000 km^2^) and large (>= 1000 km^2^) catchment data sets in our analysis.

@@

In addition, additional variables added were the latitude and longitude for the center of the watershed as an approximation of its spatial location. Using this information  reference evapotranspiration (E~0~) was extracted from the Global Aridity Index and Potential Evapo-Transpiration (ET~0~) Climate Databasev2 [@trabucco2018], if a value of E~0~ was not available from the original papers. For large watersheds, this value, similar to annual average rainfall, is only an approximation of the climate at the location.

The length of the study can be a variable influencing the change in flow [e.g. @jackson2005], as for example, more mature plantations are thought to have smaller impacts on flow. Therefore, the length of the study calculate as the difference between the starting data and completion date of the different studies was extracted from the references provided by @zhang2017.

Several additional data points from watershed studies were extracted from @zhang2011, @zhao2010, @borg1988, @thornton2007, @zhou2010, @rodriguez2010, @ruprechtetal1991 and @pena-arancibia2012, and these were checked against the existing studies to prevent overlap. In the citation column in the data set, in general the main reference for the calculated change in streamflow  was used, because sometimes the original study did not provide the quantification of the change in streamflow (i.e. Table 6 in @zhang2011).
We also removed one data point from the analysis, which corresponds to Watershed #1 (Amazon) in @zhang2017. This is because the cited reference [@roche1981] only relates to 1  and 1.5 ha paired watershed studies in French Guyana, and in which the actual change in forest cover is not recorded. Furthermore, the change in flow for watershed #76 was corrected from 600% to 157% after review of the original publication [@baker1984]. Finally, on review of all the data in @zhang2017 and @filoso2017, 29 potential duplicates were identified and flagged in the data.

@@

Similar to @zhang2017, the "dryness index" was calculated from the reference evapotranspiration and the annual average rainfall (Pa) as:
$$\tag{1}
\begin{aligned}
D = \frac{E_{0}}{Pa}
\end{aligned}$$

@@

$$\tag{2}
\begin{aligned}
\Delta Qf \% \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & \sum{X_i} + \sum{s(Z_i)} + \varepsilon
\end{aligned}$$

@@

For the model in equation 2, we initially only used the data from @zhang2017 to make sure that the additional watersheds added to the data set did not influence the results. Subsequently the analysis was repeated and the additionally identified watersheds were added.

@@

The overall dataset contains `r nrow(All_data)` observations of changes in flow, which includes the newly identified data sets and after removing identified duplicate data and lines with missing data. In contrast, the original dataset from @zhang2017 contained `r nrow(Zhang_all)` watersheds and the @filoso2017 study used 37 watersheds (Table S2 in @filoso2017). The overall distribution of changes in flow is highly skewed as is the distribution of changes in forest cover and Area. The values of changes in flow greater than 100\% and smaller than -100\% clearly create long tails on the change in flow distribution. Note also the large number of studies with 100\% forest cover reduction. Smaller watersheds dominate the database with 42\% of the data from watersheds < 1 km^2^ and 65\% of the data for watersheds < 10 km^2^.

@@

$$\tag{3}
\begin{aligned}
\Delta Qf \% \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \varepsilon
\end{aligned}$$

@@

```{r, echo =F}
pander(round(summary(linear_model_simple)$p.table,2), caption="(\\#tab:tabmodel1) Summary results of the first regression model predicting change in streamflow from change in forest cover and accounting for the direction of the change")
```

```{r, echo =F}
pander(round(summary(linear_model_simple_b)$p.table,2), caption="(\\#tab:tabmodel1b) Summary results of the first regression model predicting change in streamflow from change in forest cover and accounting for the direction of the change including the new data sets")
```

The overall variance explained in this model is not high with an adjusted *r^2^* of `r r_sq`, it generally supports the hypothesized relationship between the change in forest cover and the change in flow. The model suggests that for every 1\% change in forest cover, on the average, the flow changes 0.45\%. However the change in flow is different for forest cover decreases compared to forest cover increases. In fact, forest cover increases decrease flow by 29\% less than a similar decrease in forest cover causes flow to increase. So roughly speaking, a 1\% forest cover increase on the average decreases flow by $(1 - 0.29)*0.45\%$, while a the percentage forest cover decrease will increase flow by 0.45\%.

```{r, eval = F, echo = F}
png("residual_plot_model1.png", width = 2500, height = 1800, res = 300)
par(mfrow=c(2,2))
gam.check(linear_model_simple)
par(mfrow=c(1,1))
dev.off()
```
```{r gamcheck, echo =F, fig.cap="Residual plots for the first simple regression model indicating a slightly fat-tailed residual distribution", out.width="90%"}
include_graphics("residual_plot_model1.png")
```

Of importance here is to highlight the residuals of this regression. These are approximately normal, although there is still significant skew on the upper and lower parts of the distribution (Figure \@ref(fig:gamcheck)). In other words, the distribution of the residuals is somewhat fat-tailed. We will discuss this later.

Including the data from some of the newly identified studies indicates that this mainly strengthens the difference between the forest cover increases and decreases (Table \@ref(tab:tabmodel1) and Table \@ref(tab:tabmodel1b)), and the result indication a reduction in the mean decrease in flow as a result of forest cover change if the new data is included.

It is however it is clear from the lack of explaining power for the model, that there could be confounding factors, as alluded to in the methods. The obvious ones being watershed dryness and area (following @zhang2017):

$$\tag{4}
\begin{aligned}
\Delta Qf \% \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & Pa_{mm} + log10(Area_{km^2}) + \varepsilon
\end{aligned}$$

Where $Pa_mm$ is the annual average rainfall in mm, and a log base 10 transformation is applied to the area variable given that the distribution of Area is higly skewed (Figure 1)

<!-- At the moment using watershed rainfall, we probably want to straight away use P/E0, and explain that rainfall itself might not be a good predictor. Alternatively we can first show that Latitude is a predictor -->

```{r model2, echo = F}

linear_model2 <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + log10(Area_km2) + Pa_mm, data = All_data2)
#summary(linear_model2)
#gam.check(linear_model2)
```

```{r, echo = F}
pander(round(summary(linear_model2)$p.table,2), caption="(\\#tab:out-model2) Summary of the second model, taking into account the annual rainfall and the area of the watershed")
```

Including area and annual precipitation slightly improves the overall explaining power of the model (Table \@ref(tab:out-model2)). Annual precipitation is in fact not significant. Relative to earlier reported studies [@zhang2017; @filoso2017], the log base 10 transformed watershed area indicates a p-value of only `r round(summary(linear_model2)$p.table[4,4],2)`, suggesting a marginal impact on the change in stream flow. This supports our approach (in contrast to @zhang2017) to consider watershed area as a continuous variable and making no separation between larger and smaller watersheds. The main effects in the model remain the change in forest cover and whether this is an increase or decrease.

@@

```{r globalmap, fig.cap="Distribution of included watersheds across the globe based on reported or estimated latitude and longitude", out.width="90%", echo=F}

@@

As indicated, a further hypothesis relates to whether there is a strong spatial global gradient as captured by latitude and longitude. As the global map (Figure \@ref(fig:globalmap)) shows, the distribution of case study watersheds covers multiple continents and shows some distinct clustering in parts of the world. Of interest is whether the spatial clustering also indicates a difference in response to forest cover change:

$$\tag{5}
\begin{aligned}
\Delta Qf \% \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & Pa_{mm} + log10(Area_{km^2}) + Latitude + Longitude + \varepsilon
\end{aligned}$$

```{r model3, echo = F}
model3_includeLatLong <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + log10(Area_km2) + Pa_mm + Latitude + Longitude, data = All_data2)
#summary(model3_includeLatLong)
#gam.check(model3_includeLatLong)
```

@@

There appears to be no significant gradient in either latitude or longitude (Table \@ref(tab:out-model3)), suggesting that the distribution of the watersheds across the globe has little influence on the overall result. The total explaining power of the model is still low with an adjusted *r^2^* of `r round(summary(model3_includeLatLong)$r.sq,2)` suggesting further factors influencing the change in streamflow that are currently not included in the model.

Note that in this case the significance of the Area variable increases, and generally indicates that larger watersheds would be expected to have a lower change in streamflow, as also indicated in @zhang2017.

## Impact of the dryness index

Climate, and in particular evapotranspiration can have a significant effect on the streamflow change as represented by the dryness index, which is also highlghted by both @zhang2017 and @jackson2005. Increased evapotranspiration could lead to drier watersheds, unless balanced by rainfall (such as possibly in the tropics). This model introduces the dryness index as a linear variable and drops the annual average precipitation as a variable, as dryness is calculated from the precipitation. It also drops the Latitude and Longitude as these are indicated not to be significant.

@@

```{r model4, echo = F}
#Zhang_all2 <- Zhang_all2 %>%
#  mutate(length = To - From)
model4_dryness <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + log10(Area_km2) +  
                        Dryness, data = All_data2)
#summary(model4_dryness)
#gam.check(model4_dryness)
```

```{r, echo =F}
pander(round(summary(model4_dryness)$p.table,2), caption = "(\\#tab:out-model4) Results of the model replacing the annual precipitation with the dryness index")
```

The results from this model (Table \@ref(tab:out-model4)) confirm that dryness is a significant confounding factor related to the change in streamflow as a function of the change in forest cover change. In fact if the dryness index doubles (remembering that Dryness = 1 when E0 = Pa, so in this case E0 = 2*Pa, which is very dry), the change in runoff is ~14\% greater. However, more interesting, Latitude remains a significant predictor with each degree in latitude causing an -0.31\% change in runoff. This indicates that Dryness (i.e. an increase in radiation) alone does not explain the trend in the Latitude and some other unknown confounding factor is captured by Latitude.

```{r, echo = F}
pander(All_data2 %>% filter(Dryness > 4) %>% select(Latitude, Longitude,`Watershed name`), caption = "(\\#tab:drytable) Watersheds for which the dryness index > 4")
```

However, the result also indicates possible issues with the data, some of the Dryness values are very large (> 4) and these values have high leverage in the data. These watersheds are listed in Table \@ref(tab:drytable).

### Are some of the variables possibly non-linear?

@@

$$\tag{5}
\begin{aligned}
\Delta Qf \% \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & s(log10(Area ~(km^2)) + s(length) + s(Dryness) + \\ & s(From) + \varepsilon
\end{aligned}$$

model4b <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + From + 
                    s(log10(Area_km2), bs="ts") + s(Dryness, bs="ts") +
                 s(length, bs = "ts"),
                  data = All_data2 %>% filter(length < 60))
summary(model4b)
#gam.check(model5_all)
#plot(model5_all)
```
-->

@@

Including non-linearity increases the overall explaining power of the model to an adjusted *r^2^* of `r round(summary(model5_all)$r.sq,2)` and deviance explained of `r round(summary(model5_all)$dev.expl,2)`, but creates few changes in the significance of the variables (Table \@ref(tab:mfive-smooth)). However, it also increases the chance of over fitting, as the smoothing splines allow significant flexibility, which will be investigated later.

@@

This model explains more of the variance, but the improvement is marginal compared to the previous model with a deviance explained of `r round(summary(model6_all)$dev.expl,2)`. This indicates that the categorical variables explain a limited amount of the variance. However, interesting to note from Table \@ref(tab:msix-linear) that several of the assessment methods are significant. In particular Paired Watersheds experiments (PWE), Hydrological modelling (HM) and Statistical techniques (SH) are strongly significant ($p < 0.05$).

@@

Figure \@ref(fig:smoothsmodel6) highlights that the relationship between log10(Area km^2^) and the change in flow is essentially linear, not significant, and does not need to be smoothed (this is the value of using penalized smooths following @wood2006). It is still a negative slope, indicating that in larger watersheds the impact of changes in forest cover on streamflow is less than for smaller watersheds. Similarly, the length variable is not significant. However, both the length and Dryness variables show non-linearity, but this does not show a clear trend due to the scatter and the distributions of the data.

<!--
```{r model6_reduc, eval = F, echo = F}
model6_reduc <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
                    s(Dryness, bs="ts" )  + 
                      s(log10(Area_km2), bs="ts") +  length +
                    Assessment_technique +
                    Hydrological_regime, data = Zhang_all2)
summary(model6_reduc)

>>>>

---
title: Factors determining how catchments respond to forest cover change. Re-analysing global data sets.
author:
  - name: R. Willem Vervoort
    email: willem.vervoort@sydney.edu.au
    affiliation: DARE, The University of Sydney
    footnote: 1
  - name: Eliana Nervi
    email: eliananervi@gmail.com
    affiliation: INIA
  - name: Jimena Alonso
    email: jalonso@fing.edu.uy
    affiliation: IMFIA
address:
  - code: DARE
    address: ARC Training Centre Data Analytics for Resources and the Environment
  - code: The University of Sydney
    address: School of Life and Environmental Sciences, The University of Sydney, Sydney, NSW 2006, Australia
  - code: INIA
    address: Project Manager, FPTA 357, Instituto Nacional de Investigación Agropecuaria, INIA-Uruguay, Ruta 48 km 10, Rincon del Colorado, 90100 Canelones, Uruguay
  - code: IMFIA
    address: Institute of Fluid Mechanics and Environmental Engineering, School of Engineering, Universidad de la República, 11200 Montevideo, Uruguay
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: |
  Three recent papers review and analyse large global datasets related to impacts of forest cover on streamflow. Using three different approaches, they all find a strong relationship between forestation/de-forestation and streamflow. However, the past approaches in the literature are variable and can be substantially improved in statistical rigour, and indicate different confounding factors on the impact of forestation. The data for these three papers were reviewed, combined and re-analysed to answer the following new and older questions: 1) How is streamflow impacted by the change in forest cover as a function of catchment area; 2) how is this relationship conditioned by the length of the study, and climate; and 3) are there other possible variables that impact the observed change in streamflow? Generalised additive models were used to run flexible regressions including multiple variables. 
  Changes in forest cover cause changes in streamflow, however this change is different between deforestation and reforestation, and strongly affected by climate, with drier climates indicating larger changes in streamflow. Deforestation causes a 32\% greater change in flow compared to reforestation. Area of the catchment only affects the change in streamflow after log transformation, given the wide variety in the data from small scale paired watershed studies. Smaller studies dominate the database with 42\% of the data < 1 km2 and 65\% of the data < 10 km2. Length of the study and initial year of the study did not affect the change in flow, in contrast to other reported studies. Despite these findings, overall explained variance (38\%) is low due the quality of the inputs and additional unknown confounding factors.
journal: "Journal of Hydrology"
date: "`r Sys.Date()`"
bibliography: forest_and_water.bib
linenumbers: true
#numbersections: true
csl: elsevier-harvard.csl
output:   
  bookdown::pdf_book:
    base_format: rticles::elsevier_article
editor_options: 
  chunk_output_type: inline
---

@@

<!-- Include Filoso paper -->
For the purpose of this paper, *watershed* and *catchment* are interchangeable terms. Many of the US studies use *watershed*, while European and Australian studies use *catchment*. In particular, we retained the term "paired watershed studies" as this is the most common terminology, but further mostly use the term catchment.

@@

The conclusions of the first paper [@zhang2017] suggest that there is a distinct difference in the change in flow as a result of forestation or deforestation between small watersheds (catchments), defined as < 1000 km^2^ and large watersheds (catchments) > 1000 km^2^. While for small catchments there was no real change in runoff with changes in cover, for large catchments there was a clear trend showing a decrease in runoff with and increase in forest cover. Their main conclusion was that the response in annual runoff to forest cover was scale dependent and appeared to be more sensitive to forest cover change in water limited catchments relative to energy limited catchments [@zhang2017].

The second study [@filoso2017] was a systematic review which classified the historical research and highlighted gaps in the spatial distribution, the types of studies and the types of analysis. Their main conclusion was also that reforestation decreases streamflow, but that there were many interacting factors. For a subset of quantitative data (37) they showed a log-linear relationship between decreasing catchment size and an increasing decline in streamflow. In addition, they identified that studies with shorter periods of data collection resulted in larger declines in streamflow.

@@

The starting point of this paper is the data base of studies which were included in @zhang2017 as supplementary material. The columns in this data set are the catchment number, the	catchment name, the	Area in km^2^, the annual average precipitation	(Pa) in mm, the	forest type, hydrological regime, and climate type, the change in forest cover	in \% ($\Delta F\%$) and the change in streamflow in \% $(\Delta Qf\%)$, based on equation 1 in @zhang2017), the precipitation data type, the assessment technique, and the	source of the info, which is a citation. 
Several of these columns contain abbreviations to describe the different variables, which are summarised in Table 1.

@@

While @zhang2017 use the dryness index in their analysis, and calculate the variable climate type from this index, the potential or reference evapotranspiration was not originally included as part of the published data set. In this paper, we used only the dryness index and did not use the climate type as a variable (as they are interchangeable).
We combined the tables for small catchments (< 1000 km^2^) and large catchments (>= 1000 km^2^) in our analysis.

@@

In addition, additional variables added were the latitude and longitude for the center of the catchment as an approximation of its spatial location. Using this information  reference evapotranspiration (E~0~) was extracted from the Global Aridity Index and Potential Evapo-Transpiration (ET~0~) Climate Databasev2 [@trabucco2018], if a value of E~0~ was not available from the original papers. For large catchments, this value, similar to annual average rainfall, is only an approximation of the climate at the location.

Similar to @zhang2017, the "dryness index" was calculated from the reference evapotranspiration and the annual average rainfall (Pa) as:

\begin{equation}
D = \frac{E_{0}}{Pa} (\#eq:eq1)
\end{equation}

The length of the study can be a variable influencing the change in flow [e.g. @jackson2005; @filoso2017], as for example, more mature plantations are thought to have smaller impacts on flow or regrowth might follow a "Kuczera curve" [@kuczera1987]. Therefore, the length of the study calculate as the difference between the starting data and completion date of the different studies was extracted from the references provided by @zhang2017. The length of the study was already included in the data from @filoso2017, but these were checked against the original publications.

Several additional data points from catchment studies were extracted from @zhang2011, @zhao2010, @borg1988, @thornton2007, @zhou2010, @rodriguez2010, @ruprechtetal1991 and @pena-arancibia2012, and these were checked against the existing studies to prevent overlap. In the citation column in the data set, in general the main reference for the calculated change in streamflow  was used, because sometimes the original study did not provide the quantification of the change in streamflow (i.e. Table 6 in @zhang2011).
We also removed one data point from the analysis, which corresponds to catchment #1 (Amazon) in @zhang2017. This is because the cited reference [@roche1981] only relates to 1  and 1.5 ha paired catchment studies in French Guyana, and in which the actual change in forest cover is not recorded. Furthermore, the change in flow for catchment #76 was corrected from 600% to 157% after review of the original publication [@baker1984]. Finally, on review of all the data in @zhang2017 and @filoso2017, 29 potential duplicates were identified and flagged in the data.

@@

@@

\begin{align}
\Delta Qf \% \sim &~ \Delta \% forest~cover_{positive} + sign_{forest~cover} + \notag \\ 
& \sum{X_i} + \sum{s(Z_i)} + \varepsilon (\#eq:eq2)
\end{align}

@@

For the model in equation 2, we initially only used the data from @zhang2017 to make sure that the additional catchments added to the data set did not influence the results. Subsequently the analysis was repeated and the additionally identified catchments were added.

@@

The overall dataset contains `r nrow(All_data)` observations of changes in flow, which includes the newly identified data sets and after removing identified duplicate data and lines with missing data. In contrast, the original dataset from @zhang2017 contained `r nrow(Zhang_all)` catchments and the @filoso2017 study used 37 catchments (Table S2 in @filoso2017). The overall distribution of changes in flow is highly skewed as is the distribution of changes in forest cover and Area. The values of changes in flow greater than 100\% and smaller than -100\% clearly create long tails on the change in flow distribution. Note also the large number of studies with 100\% forest cover reduction. Smaller catchments dominate the database with 42\% of the data from catchments < 1 km^2^ and 65\% of the data for catchments < 10 km^2^.

@@

\begin{equation}
\Delta Qf \% \sim ~ \Delta \% forest~cover_{positive} + sign_{forest~cover} + \varepsilon (\#eq:eq3)
\end{equation}

@@

```{r, echo =F}
pander(round(rbind(summary(linear_model_simple)$p.table,
                       summary(linear_model_simple_b)$p.table),2), caption="(\\#tab:tabmodel1) Summary results of the first regression model predicting change in streamflow from change in forest cover and accounting for the direction of the change. The first three rows relate to the model using the original data base from Zhang et al. (2017). The bottom three rows are the results of the model including the new data")
```

The overall variance explained in this model (equation \@ref(eq:eq3)) is not high with an adjusted *r^2^* of `r r_sq`, it generally supports the hypothesized relationship between the change in forest cover and the change in flow. The model suggests that for every 1\% change in forest cover, on the average, the flow changes 0.45\%. However the change in flow is different for forest cover decreases compared to forest cover increases. In fact, forest cover increases decrease flow by 29\% less than a similar decrease in forest cover causes flow to increase. So roughly speaking, a 1\% forest cover increase on the average decreases flow by $(1 - 0.29)*0.45\%$, while a the percentage forest cover decrease will increase flow by 0.45\%.

```{r, eval = F, echo = F}
png("residual_plot_model1.png", width = 2500, height = 1800, res = 300)
par(mfrow=c(2,2))
gam.check(linear_model_simple)
par(mfrow=c(1,1))
dev.off()
```

```{r gamcheck, echo =F, fig.cap="Residual plots for the first simple regression model indicating a slightly fat-tailed residual distribution", out.width="90%"}
include_graphics("residual_plot_model1.png")
```

Of importance here is to highlight the residuals of this regression (equation \@ref(eq:eq3) and Figure \@ref(fig:gamcheck)). These are approximately normal, although there is still significant skew on the upper and lower parts of the distribution (Figure \@ref(fig:gamcheck)). In other words, the distribution of the residuals is somewhat fat-tailed. We will discuss this later.

Including the data from some of the newly identified studies indicates that this mainly strengthens the difference between the forest cover increases and decreases (Table \@ref(tab:tabmodel1)), and the result indicate a reduction in the mean decrease in flow as a result of forest cover change if the new data is included.

It is however it is clear from the lack of explaining power for the model, that there could be confounding factors, as alluded to in the methods. The obvious ones being catchment dryness and area (following @zhang2017), which we will analyse later.

@@

```{r globalmap, fig.cap="Distribution of included catchments across the globe based on reported or estimated latitude and longitude", out.width="90%", echo=F}

@@

As indicated, an initial hypothesis relates to whether there is a strong spatial global gradient as captured by latitude and longitude. These data were added for the different studies, mostly by using the data reported by the authors, but in some cases approximating the location of the centre of the catchment using Google Maps. In the dataset, an additional column is added to indicate the source of the location data.  As the global map (Figure \@ref(fig:globalmap)) shows, the distribution of case study catchments covers multiple continents and shows some distinct clustering in parts of the world. Of interest is whether the spatial clustering also indicates a difference in response to forest cover change:

\begin{align}
\Delta Qf \% \sim ~ & \Delta \% forest~cover_{positive} + sign_{forest~cover} + \notag \\ &  Latitude + Longitude + \varepsilon (\#eq:eq4)
\end{align}

```{r model3, echo = F}
model3_includeLatLong <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + Latitude + Longitude, data = All_data2)
#summary(model3_includeLatLong)
#gam.check(model3_includeLatLong)
```

@@

There appears to be no significant gradient in either latitude or longitude (Table \@ref(tab:out-model3)), suggesting that the distribution of the catchments across the globe has little influence on the overall result. The total explaining power of the model is still low with an adjusted *r^2^* of `r round(summary(model3_includeLatLong)$r.sq,2)` suggesting further factors influencing the change in streamflow that are currently not included in the model.

## Impact of climate

Climate, and in particular evapotranspiration can have a significant effect on the streamflow change as represented by the dryness index, which is also highlghted by both @zhang2017 and @jackson2005. Increased evapotranspiration could lead to drier catchments, unless balanced by rainfall (such as possibly in the tropics). Initially, we tested models using annual average precipitation (*Pa (mm)*), but because of the interactions between precipitation, evapotranspiraton and the dryness index, we concentrated on the dryness index as the key variable. Given that Latitude and Longitude were not significant, we dropped these from the model.

\begin{align}
\Delta Qf \% \sim ~ & \Delta \% forest~cover_{positive} + sign_{forest~cover} + \notag \\ & Dryness + \varepsilon (\#eq:eq5)
\end{align}

@@

```{r model4, echo = F}
#Zhang_all2 <- Zhang_all2 %>%
#  mutate(length = To - From)
model4_dryness <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign +  
                        Dryness, data = All_data2)
#summary(model4_dryness)
#gam.check(model4_dryness)
```

```{r, echo =F}
pander(round(summary(model4_dryness)$p.table,2), caption = "(\\#tab:out-model4) Results of the model including the dryness index")
```

The results from this model (equation \@ref(eq:eq5) and Table \@ref(tab:out-model4)) confirm that dryness is a clear confounding factor related to the change in streamflow as a function of the change in forest cover change. In this case the evidence is not very strong (p = `r round(summary(model4_dryness)$p.table[4,4],2)`. However, if the dryness index doubles (remembering that Dryness = 1 when E0 = Pa, so in this case E0 = 2×Pa, which is very dry), the change in runoff is ~6\% greater. Again, overall variance explained is not very much, with an adjusted *r^2^* of `r round(summary(model4_dryness)$r.sq,2)`.

```{r, echo = F}
pander(All_data2 %>% filter(Dryness > 4) %>% select(Latitude, Longitude,`Watershed name`), caption = "(\\#tab:drytable) catchments for which the dryness index > 4")
```

There possible issues with the data, a few of the catchments have Dryness values that are very large (> 4) and these values have high leverage in the data, affecting the residual distribution. These catchments are listed in Table \@ref(tab:drytable).

## Is there a distinct effect of area?

The major hypothesis to test is the effect of area on the change in flow, following the analysis by @zhang2017 and @filoso2017. Given the highly skewed distribution of the catchment areas (Figure \@ref(fig:datagraphs)), a log base 10 transformation was applied to the variable *Area (km^2^)*.

\begin{align}
\Delta Qf \% \sim ~ & \Delta \% forest~cover_{positive} + sign_{forest~cover} + \notag \\ & log10(Area ~(km^2)) + Dryness + \varepsilon (\#eq:eq6)
\end{align}

```{r modelArea, echo = F}
#Zhang_all2 <- Zhang_all2 %>%
#  mutate(length = To - From)
model_Area <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign +  
                        Dryness + log10(Area_km2), data = All_data2)
#summary(model_Area)
#gam.check(model_Area)
```

```{r, echo =F}
pander(round(summary(model_Area)$p.table,2), caption = "(\\#tab:out-modelArea) Results of the model including Area and the dryness index")
```

The results of this model (Equation \@ref(eq:eq6)) clearly indicate a reduction in the effect of forest cover change with Area (km^2^) (Table \@ref(tab:out-modelArea)). In fact, the results suggests that for every additional 10 km^2^ in catchment size the mean change in flow reduces by 3.5\%. Another interesting fact to note is that with the inclusion of Area (km^2^) as a variable in the model, the effect of Dryness becomes more significant, possibly suggesting an interaction between Dryness and Area. Including this interaction suggest that the interaction term (log10(Area) by Dryness) would be significant, but this replaces the effect of Area (results not shown).

```{r modelArea_b, echo = F, eval =F}
#Zhang_all2 <- Zhang_all2 %>%
#  mutate(length = To - From)
model_Area_b <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign +  
                        Dryness*log10(Area_km2), data = All_data2)
summary(model_Area_b)
#gam.check(model_Area)
```

## Are some of the variables possibly non-linear?

@@

\begin{align}
\Delta Qf \% \sim &~ \Delta \% forest~cover_{positive} + sign_{forest~cover} + \notag \\ & s(log10(Area ~(km^2)) + s(length) + s(Dryness) + \\ & s(From) + \varepsilon (\#eq:eq7)
\end{align}

@@

Including non-linearity (Equation \@ref(eq:eq7)) increases the overall explaining power of the model to an adjusted *r^2^* of `r round(summary(model5_all)$r.sq,2)` and deviance explained of `r round(summary(model5_all)$dev.expl,2)`, but creates a few changes in the significance of the variables (Table \@ref(tab:mfive-smooth)). For example, the smoothed variable for Area (km^2^) is no longer a strong effect explaining variations in changes in stream flow. Clearly *length* is also not explaining the variation. In contrast, *From*, which indicate the start date of the study (and therefore the age of the study) is a significant effect.  
However, it also increases the chance of over fitting, as the smoothing splines allow significant flexibility, which will be investigated later. Including interactions into the smooths is possible, but the results are even more difficult to interpret given the high flexibility of the two-dimensional smooth.

@@

This model explains more of the variance, but the improvement is marginal compared to the previous model with a deviance explained of `r round(summary(model6_all)$dev.expl,2)`. This indicates that the categorical variables explain a limited amount of the variance. However, interesting to note from Table \@ref(tab:msix-linear) that several of the assessment methods are significant. In particular Paired Watersheds experiments (PWE), Hydrological modelling (HM) and Statistical techniques (SH) are strongly significant ($p < 0.05$). 
In this case, *Area (km^2)* is no longer a significant predictor.

@@

Figure \@ref(fig:smoothsmodel6) highlights that the relationship between log10(Area km^2^) and the change in flow is essentially linear, not significant, and does not need to be smoothed (this is the value of using penalized smooths following @wood2006). It still has a negative slope, indicating that in larger catchments the impact of changes in forest cover on streamflow is less than for smaller catchments. Similarly, the length variable is not significant. However, both the length and Dryness variables show non-linearity, but this does not show a clear trend due to the scatter and the distributions of the data.
